\documentclass{ximera}
%% handout
%% nohints
%% space
%% newpage
%% numbers

\input{preamble.tex} %% we can turn off input when making a master document

\prerequisites{none}
\outcomes{ximeraLatex}

\title{How to compute percentages.}

\begin{document}
\begin{abstract}
We review the concept of percentages and introducte \protect\url{ConsumerMath.org}.
\end{abstract}
\maketitle


The word \emph{percent} means ``for each hundred.'' For example, saying Joseph earns $60\%$ as much as Elaine is saying he earns $\$60$ for each $\$100$ she earns.

To calculate a percentage, the relationship between the amount that constitutes the whole, called the \emph{reference amount} and the amount under consideration must be found. This is done via the following formula:

\begin{equation}\label{basicpercentage}
\text{Percentage}=\frac{\text{Part}}{\text{Whole}}\times 100
\end{equation}

As an example,
\begin{equation*}
\frac{15}{25}\times 100=60
\end{equation*}
means that $15$ is $60\%$ of $25$.

The relationship in equation (\ref{basicpercentage}) can be used to find any of the three pieces -- the percentage, the part, the whole -- if the other two are known by using algebra to solve for the missing piece.

As an example, to find $70\%$ of $220$
\begin{align*}
70&=\frac{\text{Part}}{220}\times 100\\
70\div 100 &=\frac{\text{Part}}{220}\\
0.70\times 220 &=\text{Part}\\
154&=\text{Part}
\end{align*}
and so $70\%$ of $220$ is $154$.

\newcommand\customfont[1]{{\usefont{T1}{custom}{m}{n} #1 }}

\begin{question}
The label on a 12 oz.\ can of \customfont{Coca-Cola}\ states that it contains 39 grams of carbohydrates, accounting for $13\%$ of the recommended daily value. How many grams of carbohydrates are recommended daily?

\begin{solution}
\begin{hint}
\begin{equation*}13=\frac{39}{\emph{\text{Whole}}}\times 100\end{equation*}
\end{hint}
\begin{expression-answer}
function validator(p) {
    if (p == 300)
      return 1;
    if (p <> 300)
      return 0;
  }
\end{expression-answer}
\end{solution}

Nice work!
\end{question}

\begin{question}
For fiscal years 2013--2015, the Nevada legislature appropriated $9.0\%$ of its $\$20.1$ billion budget to infrastructre. How much was appropriated for infrastructure? 	

\begin{solution}
\begin{hint}
\begin{equation*}9=\frac{\emph{\text{Part}}}{\$20.1\emph{\text{ billion}}}\times 100\end{equation*}
\end{hint}
\begin{hint}
If you use $\$20,100,000,000$ in your calculations, your answer will be in dollars. If you use $\$20.1$ in your calculations, your answer will be in billions of dollars.
\end{hint}
\begin{expression-answer}
function validator(p) {
    if (p - 1.809) &lt; 0.01)
      return 1;
    if (p - 1809000000 &lt; 10000000)
      return 1;
      return 0;
  }
\end{expression-answer}
\end{solution}	

Nice work!
\end{question}

\begin{question}
If my taxi fare is $\$25.50$ and I pay the driver $\$30$, what percent tip am I giving?
  \begin{solution}
    \begin{multiple-choice}
    	\choice{$85\%$}
        \choice{$20\%$}
        \choice[correct]{$17.6\%$}
        \choice{$15\%$}
        \choice{$1.2\%$}
    \end{multiple-choice}
    \begin{hint}
    What is the \emph{amount} of the tip? This is your \emph{part}.
    \end{hint}
    \begin{hint}
   What is the amount of the basic fare? This is your \emph{whole}.
    \end{hint}
  	\begin{hint}
    \begin{equation*}\emph{\text{Percentage}}=\frac{\emph{\text{Part}}}{\emph{\text{Whole}}}\times 100\end{equation*}
    \end{hint}
  \end{solution}
\end{question}

A significant piece of the Consumer Mathematics portion of this course involves the use of a financial simulator at \url{ConsumerMath.org}.


\end{document}