\documentclass{ximera}
%% handout
%% nohints
%% space
%% newpage
%% numbers

\input{preamble.tex} %% we can turn off input when making a master document

\prerequisites{none}
\outcomes{ximeraLatex}

\title{What is a Function?}

\begin{document}
\begin{abstract}
We define functions and see some examples.
\end{abstract}
\maketitle


If two variables, $x$ and $y$, are related to each other, we say that $y$ is a \emph{function} of $x$ if each $x$ value is paired with only one $y$-value. In other words, each input value, $x$, results in exactly one output value, $y$.

For example, if $t$ is the time of day, and $s$ is the speed of your car, then is $s$ a function of $t$? At any time of day there is only one speed your car is traveling, so $s$ is a function of $t$. In other words $s$ depends on $t$.

Is $t$ a function of $s$? No. There will almost certainly be some some speed that your car reached at two different times. Thus $t$ is not a function of $s$. The speed could be determined by knowing the time, but the time could not be determined just by knowing the speed.

Another example can be found at a vending machine. Let $x$ be the number of candy bars you buy (for a dollar) and $y$ be the amount you spend on candy bars. Then $y$ is a function of $x$ because the number of candy bars determines the exact amount you spend. Also, $x$ is a function of $y$ because the number of candy bars can be determined by the amount you spent.


\begin{question}
Select each option below where the first item is a function of the second item.
  \begin{solution}
    \begin{multiple-choice}
      \choice[correct]{---the amount on a paycheck and the number of hours worked at an hourly wage job}
      \choice{---a man's height and his father's height}
      \choice{---the number of calories in a meal and the amount you pay for the meal}
      \choice[correct]{---the area of a square and the length of its sides}
      \choice[correct]{---the length of the sides of a square and its area}
      \choice{---the area of a rectangle and its perimeter}
      \choice{---driverse licence numbers and drivers}
    \end{multiple-choice}
    \begin{hint}
    If the hourly wage is set, then we should only expect one possible payout.
    \end{hint}
    \begin{hint}
    Is it possible for two father's of the same height to have sons of different heights?
    \end{hint}
    \begin{hint}
    Is it possible that two meals that share the same price might have a different number of calories?
    \end{hint}
    \begin{hint}
    Does the formula for the area of a square depend on the side length?
    \end{hint}
    \begin{hint}
    The length of the side of a square is always the square root of the area
    \end{hint}
    \begin{hint}
    A $3\times 5$ rectangle has the same perimeter as a $2\times 6$ rectangle. Do they have the same area?
    \end{hint}
    \begin{hint}
    Does any driver have more than one driver's license numbers?
    \end{hint}
  \end{solution}
\end{question}




\end{document}